### 2024-06-11 products/manuscript/Manuscript.qmd
Updated Manuscript

---
title: "Manuscript for a Data Analysis Project: Diabetes Diagnosis"
subtitle: ""
author: Austin Thrash
date: today
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
    highlight-style: github
bibliography: ../dataanalysis-template-references.bib
csl: ../apa.csl
---

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

# Summary/Abstract

*\*Summary will be finalized in later part of project and will include setup and results*

{{< pagebreak >}}

# Introduction

## General Background Information

*\*more background information will be added later, I will describe what diabetes is, what are the notable influences on diagnosis, and what the dataset provides in relation to other studies on diabetes diagnosis.*

## Description of data and data source

This dataset comes from UC Irvine Machine Learning Repository. This data has been gathered and processed from the CDCs website in 2014. This dataset includes over 250,000 instances (observations) and 21 features, ranging from personal information (income/education) to health history (drinking/smoking/physical health). Bellow are the features and there respective descriptions

| Feature              | Description                                                                                                                                                                                                                                                                                                                             |
|----------------------|-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| ID                   | Patient ID                                                                                                                                                                                                                                                                                                                              |
| Diabetes_binary      | 0 = no diabetes 1 = prediabetes or diabetes                                                                                                                                                                                                                                                                                             |
| HighBP               | 0 = no high BP 1 = high BP                                                                                                                                                                                                                                                                                                              |
| HighChol             | 0 = no high cholesterol 1 = high cholesterol                                                                                                                                                                                                                                                                                            |
| CholCheck            | 0 = no cholesterol check in 5 years 1 = yes cholesterol check in 5 years                                                                                                                                                                                                                                                                |
| BMI                  | Body Mass Index                                                                                                                                                                                                                                                                                                                         |
| Smoker               | Have you smoked at least 100 cigarettes in your entire life? \[Note: 5 packs = 100 cigarettes\] 0 = no 1 = yes                                                                                                                                                                                                                          |
| Stoke                | (Ever told) you had a stroke. 0 = no 1 = yes                                                                                                                                                                                                                                                                                            |
| HeartDiseaseorAttack | coronary heart disease (CHD) or myocardial infarction (MI) 0 = no 1 = yes                                                                                                                                                                                                                                                               |
| PhysActivity         | physical activity in past 30 days - not including job 0 = no 1 = yes                                                                                                                                                                                                                                                                    |
| Fruits               | Consume Fruit 1 or more times per day 0 = no 1 = yes                                                                                                                                                                                                                                                                                    |
| Veggies              | Consume Vegetables 1 or more times per day 0 = no 1 = yes                                                                                                                                                                                                                                                                               |
| HvyAlchoholConsump   | Heavy drinkers (adult men having more than 14 drinks per week and adult women having more than 7 drinks per week) 0 = no 1 = yes                                                                                                                                                                                                        |
| AnyHealthcare        | Have any kind of health care coverage, including health insurance, prepaid plans such as HMO, etc. 0 = no 1 = yes                                                                                                                                                                                                                       |
| NoDocbcCost          | Was there a time in the past 12 months when you needed to see a doctor but could not because of cost? 0 = no 1 = yes                                                                                                                                                                                                                    |
| GenHlth              | Would you say that in general your health is: scale 1-5 1 = excellent 2 = very good 3 = good 4 = fair 5 = poor                                                                                                                                                                                                                          |
| MentHlth             | Now thinking about your mental health, which includes stress, depression, and problems with emotions, for how many days during the past 30 days was your mental health not good? scale 1-30 days                                                                                                                                        |
| PhysHlth             | Now thinking about your physical health, which includes physical illness and injury, for how many days during the past 30 days was your physical health not good? scale 1-30 days                                                                                                                                                       |
| DiffWalk             | Do you have serious difficulty walking or climbing stairs? 0 = no 1 = yes                                                                                                                                                                                                                                                               |
| Sex                  | 0 = female 1 = male                                                                                                                                                                                                                                                                                                                     |
| Age                  | 13-level age category (\_AGEG5YR see codebook) 1 = 18-24 9 = 60-64 13 = 80 or older                                                                                                                                                                                                                                                     |
| Education            | Education level (EDUCA see codebook) scale 1-6 1 = Never attended school or only kindergarten 2 = Grades 1 through 8 (Elementary) 3 = Grades 9 through 11 (Some high school) 4 = Grade 12 or GED (High school graduate) 5 = College 1 year to 3 years (Some college or technical school) 6 = College 4 years or more (College graduate) |
| Income               | Income scale (INCOME2 see codebook) scale 1-8 1 = less than \$10,000 5 = less than \$35,000 8 = \$75,000 or more                                                                                                                                                                                                                        |

<https://archive.ics.uci.edu/dataset/891/cdc+diabetes+health+indicators>

## Questions/Hypotheses to be addressed

"What are the key health indicators that predict the likelihood of an individual being diagnosed with diabetes?" Specifically, identifying which factors are most strongly associated with diabetes prevalence in the population.

{{< pagebreak >}}

# Methods

*\*Analysis methods will be included in later part of project*

```{r}
library(readr)
```

## Data acquisition

### Explain the question to be answered using the data

The main question I would like to answer with this project is "What are the key health indicators that predict the likelihood of an individual being diagnosed with diabetes?".

### What will be your outcome(s) of interest (if any)?

The primary outcome of interest will be whether an individual has been diagnosed with diabetes or not. This outcome will be represented as a binary variable where 1 indicates a diabetes diagnosis and 0 indicates no diabetes.

### What (if any) specific predictors will you focus on?

The specific predictors I will focus on are age, gender, BMI, consumption of fruits and vegetables, and frequency of exercise. Although these are the main predictors I will focus on, I'll also use data analytic techniques to identify which predictors are of importance or have relationships with the response variable.

### What relations/patterns are you looking for in the data?

I will look for patterns ad relationships between the predictors and the outcome of diabetes diagnosis. More specifically, I will analyze correlations between individual predictors and the response variable, interaction effects between predictors and the response variable, and predictive models such as logistic regression, decision trees, and machine learning classifiers.

## Data import and cleaning

*\*To be added in later portion of project*

## Statistical analysis

For my analysis, I will first look at the distributions of each variable to identify if the dataset is balanced and outliers. I do not think any data cleaning will need to be done as that was completed when the data was collected from the CDCs website. Next, I will create a correlation matrix to view the correlations between each predictor and the response variable. After viewing the correlations, I will then use simple linear regression and random forest models to identify variables of importance, these variables will be noted and used later during the modeling phase. I will then split the data set into a training and test set using a 80/20 split and begin the modeling phase. This section will be filled out more once we are further into the class and have discussed more approaches/methods.

***\* Any section past results has not been modified yet. I only completed the sections required for the first part.***

{{< pagebreak >}}

# Results

## Exploratory/Descriptive analysis

***\*To be added in later portion of project***

*Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project.*

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below. I generally recommend the `here` package.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS("../../results/tables/summarytable.rds")
knitr::kable(resulttable)
```

## Basic statistical analysis

***\*To be added in later portion of project***

*To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
```

## Full analysis

***\*To be added in later portion of project***

*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here.*

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
knitr::kable(resulttable2)
```

{{< pagebreak >}}

# Discussion

## Summary and Interpretation

***\*To be added in later portion of project***

*Summarize what you did, what you found and what it means.*

## Strengths and Limitations

***\*To be added in later portion of project***

*Discuss what you perceive as strengths and limitations of your analysis.*

## Conclusions

***\*To be added in later portion of project***

*What are the main take-home messages?*

*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like.

{{< pagebreak >}}

# References

***\*To be added in later portion of project***

### 2024-06-11 products/manuscript/Manuscript.qmd
Fix Links

@@
     highlight-style: github
-bibliography: ../dataanalysis-template-references.bib
-csl: ../apa.csl
+bibliography: ../../assets/dataanalysis-template-references.bib
+csl: ../../assets/apa.csl
 ---

### 2024-06-15 code/eda-code/EDA-Code.qmd
Completed Final Project - Part 2
Rewrote/reformatted some wording in the manuscript. Cleaned and explored the data, creating plots of distributions and correlations, and then added them to the manuscript. Removed files that did not go with my project/from the template. Made sure it was reproducible and was able to knit into an html.

---
title: "EDA Code"
subtitle: ""
author: Austin Thrash
date: today
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
    highlight-style: github
csl: ../../assets/apa.csl
---

```{r}
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
library(readr)
library(tidyr) # data wrangling
library(rpart) # DT
library(tidyverse)
library(car)
library(tree)
library(caret)
library(MASS)
library(ROCR)
library(randomForest)
library(corrplot)
library(ggcorrplot) # corr plot
library(reshape2) # heat map
library(GGally) # Pais ggplot2
```

# Reading in the Data

```{r}
#Path to data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","processeddata.rds")
#load data
mydata <- readRDS(data_location)
```

```{r}
summary_df = skimr::skim(mydata)
print(summary_df)
# save to file
summarytable_file = here("results", "tables", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```

# Simple Exploration

```{r}

```

# Plots of Response/Features

## Histogram

```{r}
p1 <- mydata %>% ggplot(aes(x=Diabetes_012)) + geom_histogram() 
plot(p1)
figure_file = here("results", "figures","diabetes-distribution.png")
ggsave(filename = figure_file, plot=p1) 
```

```{r}
p2 <- mydata %>% ggplot(aes(x=Smoker)) + geom_histogram() 
plot(p2)
figure_file = here("results", "figures","smoker-distribution.png")
ggsave(filename = figure_file, plot=p2) 
```

```{r}
p3 <- mydata %>% ggplot(aes(x=BMI)) + geom_histogram() 
plot(p3)
figure_file = here("results", "figures","BMI-distribution.png")
ggsave(filename = figure_file, plot=p3) 
```

```{r}
p4 <- mydata %>% ggplot(aes(x=PhysHlth)) + geom_histogram() 
plot(p4)
figure_file = here("results", "figures","PhysHlth-distribution.png")
ggsave(filename = figure_file, plot=p4) 
```

```{r}
p5 <- mydata %>% ggplot(aes(x=Age)) + geom_histogram() 
plot(p5)
figure_file = here("results", "figures","Age-distribution.png")
ggsave(filename = figure_file, plot=p5) 
```

## Correlation Matrix

```{r}
library(corrplot)

# Compute the correlation matrix for numerical attributes
numerical_data <- mydata[, sapply(mydata, is.numeric)]
cor_matrix <- cor(numerical_data)  # Handles missing values by excluding them

# Define the file path to save the plot
figure_file <- here("results", "figures", "correlation_matrix.png")

# Save the correlation matrix plot
png(filename = figure_file, width = 800, height = 600)
corrplot(cor_matrix, method = "color", type = "upper", order = "hclust",
         tl.col = "black", tl.srt = 45, addCoef.col = "black", number.cex = 0.5)
dev.off()
```

```{r}
mydata$Diabetes_012 <- as.factor(mydata$Diabetes_012)
```

# Features of Importance

## Linear Model

```{r}
glm.full = glm(Diabetes_012 ~ ., data = mydata, family = binomial)

glm.vif = vif(glm.full)

glm.vif
```

## Random Forest

```{r}
set.seed(12345)
import_features = train(Diabetes_012 ~ ., 
               data = mydata, 
               method = "rf", 
               importance = TRUE, 
               trControl = trainControl(method = "cv", number = 2, verbose = TRUE))
```

```{r}
varimp_plot <- plot(varImp(import_features))

# Define the file path to save the plot
figure_file <- here("results", "figures", "variable_importance.png")

# Save the plot
png(filename = figure_file, width = 1000, height = 800)
print(varimp_plot)
dev.off()
```

### 2024-06-15 code/processing-code/processingfile-v1.qmd
Completed Final Project - Part 2
Rewrote/reformatted some wording in the manuscript. Cleaned and explored the data, creating plots of distributions and correlations, and then added them to the manuscript. Removed files that did not go with my project/from the template. Made sure it was reproducible and was able to knit into an html.

---
title: "An example cleaning script"
author: "Andreas Handel"
date: "2024-02-07"
output: html_document
---

# Processing script

This contains the same code and comments/information as `processingcode.R`.

This just shows it as Quarto file, to give you an idea how to do it in a setup that combines code and text in a single file.

See the other Quarto file for my currently preferred approach of pulling code from the R script into the Quarto file.

# Setup

Load needed packages. make sure they are installed.

```{r}
library(readxl) #for loading Excel files
library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
```

# Data loading

Note that for functions that come from specific packages (instead of base R), I often specify both package and function like so:
package::function() that's not required one could just call the function specifying the package makes it clearer where the function "lives",
but it adds typing. You can do it either way.

```{r}
# path to data
# note the use of the here() package and not absolute paths
data_location <- here::here("data","raw-data","diabetes.csv")
rawdata <- read.csv(data_location)
```

# Check data

Several ways of looking at the data

```{r}
dplyr::glimpse(rawdata)
summary(rawdata)
head(rawdata)
skimr::skim(rawdata)
```

# Cleaning

This data set was cleaned prior to upload online, from reviewing the data, it does not seem any further cleaning or processing is required.

# Save data

```{r}
save_data_location <- here::here("data","processed-data","processeddata.rds")
saveRDS(rawdata, file = save_data_location)
```

# Notes

Removing anyone who had "faulty" or missing data is one approach. It's often not the best. based on your question and your analysis approach, you might want to do cleaning differently (e.g. keep individuals with some missing information).

### 2024-06-15 products/manuscript/Manuscript.qmd
Completed Final Project - Part 2
Rewrote/reformatted some wording in the manuscript. Cleaned and explored the data, creating plots of distributions and correlations, and then added them to the manuscript. Removed files that did not go with my project/from the template. Made sure it was reproducible and was able to knit into an html.

--- 2024-06-11 products/manuscript/Manuscript.qmd
@@
     highlight-style: github
-bibliography: ../dataanalysis-template-references.bib
-csl: ../apa.csl
+csl: ../../assets/apa.csl
 ---
@@
 | Feature              | Description                                                                                                                                                                                                                                                                                                                             |
-|----------------------|-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
+|------------|------------------------------------------------------------|
 | ID                   | Patient ID                                                                                                                                                                                                                                                                                                                              |
@@
 
-***\*To be added in later portion of project***
-
-*Use a combination of text/tables/figures to explore and describe your data.
-Show the most important descriptive results here.
-Additional ones should go in the supplement.
-Even more can be in the R and Quarto files that are part of your project.*
-
-@tbl-summarytable shows a summary of the data.
-
-Note the loading of the data providing a **relative** path using the `../../` notation.
-(Two dots means a folder up).
-You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path.
-You can also use the `here` R package to create paths.
-See examples of that below.
-I generally recommend the `here` package.
+### Diabetes Distribution
 
 ```{r}
-#| label: tbl-summarytable
-#| tbl-cap: "Data summary table."
-#| echo: FALSE
-resulttable=readRDS("../../results/tables/summarytable.rds")
-knitr::kable(resulttable)
+knitr::include_graphics(here("results","figures","diabetes-distribution.png"))
 ```
+
+### BMI Distribution
+
+```{r}
+knitr::include_graphics(here("results","figures","BMI-distribution.png"))
+```
+
+### PhysHlth Distribution
+
+```{r}
+knitr::include_graphics(here("results","figures","PhysHlth-distribution.png"))
+```
+
+### Age Distribution
+
+```{r}
+knitr::include_graphics(here("results","figures","Age-distribution.png"))
+```
+
+### Smoker Distribution
+
+```{r}
+knitr::include_graphics(here("results","figures","smoker-distribution.png"))
+```
+
+### Correlation Matrix
+
+```{r}
+knitr::include_graphics(here("results","figures","correlation_matrix.png"))
+```
+
+### Feature Importance - Linear Regression
+
+\*Will fix and put into readable table
+
+```         
+            HighBP             HighChol            CholCheck                  BMI               Smoker               Stroke HeartDiseaseorAttack 
+            1.141777             1.070725             1.011236             1.124242             1.073702             1.071638             1.150480 
+        PhysActivity               Fruits              Veggies    HvyAlcoholConsump        AnyHealthcare          NoDocbcCost              GenHlth 
+            1.140832             1.109908             1.105040             1.013591             1.100212             1.157580             1.743109 
+            MentHlth             PhysHlth             DiffWalk                  Sex                  Age            Education               Income 
+            1.281663             1.697022             1.499052             1.109269             1.326329             1.299898             1.505502 
+```
+
+### Feature Importance - Random Forest
+
+```{r}
+knitr::include_graphics(here("results","figures","variable_importance.png"))
+```
+
+Overall, the variables do not show much correlation with one another, however I have a feeling that I may need to do more data processing, i.e.
+making categorical variables into factors and removing them from the correlation matrix.
+I decided to include some distributions of variables that I found to be interesting, such as BMI, Age, PhysHlth, and Smoker.
+After running both a linear and random forest model to determine the most important features, it looks as if BMI, HighBP, HeartDiseaseorAttack, and HvyAlchoholConsump are the most important features to include in the models.
 
@@
 
-*To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*
-
-@fig-result shows a scatterplot figure produced by one of the R scripts.
-
-```{r}
-#| label: fig-result
-#| fig-cap: "Height and weight stratified by gender."
-#| echo: FALSE
-knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
-```
-
 ## Full analysis
@@
 ***\*To be added in later portion of project***
-
-*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here.*
-
-Example @tbl-resulttable2 shows a summary of a linear model fit.
-
-```{r}
-#| label: tbl-resulttable2
-#| tbl-cap: "Linear model fit table."
-#| echo: FALSE
-resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
-knitr::kable(resulttable2)
-```
 
@@
 
-*Summarize what you did, what you found and what it means.*
-
 ## Strengths and Limitations
@@
 
-*Discuss what you perceive as strengths and limitations of your analysis.*
-
 ## Conclusions
@@
 ***\*To be added in later portion of project***
-
-*What are the main take-home messages?*
-
-*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*
-
-This paper [@leek2015] discusses types of analyses.
-
-These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.
-
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like.
 

### 2024-06-22 code/analysis-code/statistical_analysis_code.qmd
Updated Statistical Analysis Code - Final Project Part 3
Worked on the statistical analysis code for part 3 of the final project.

Updated manuscript with new relevant information from statistical analysis process.

Updated EDA code to create histograms all in one image instead of seperate images.

---
title: "Manuscript for a Data Analysis Project: Diabetes Diagnosis"
subtitle: ""
author: Austin Thrash
date: today
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
    highlight-style: github
csl: ../../assets/apa.csl
---

## SETUP - Loading In Packages

```{r}
# Loading in necessary libraries
library(ISLR2)
library(MASS)
library(boot)
library(glmnet)
library(ggplot2) 
library(pls)
library(leaps)
library(gam)
library(fastDummies)
library(caret)
library(nnet)
library(MLmetrics)
library(randomForest)
```

## SETUP - Loading In The Data

```{r}
# This chunk is used to load in the data and display its structure:
# ------ OUTPUT: ------
# 'data.frame':	253680 obs. of  22 variables:
# $ Diabetes_012        : num  0 0 0 0 0 0 0 0 2 0 ...
# $ HighBP              : num  1 0 1 1 1 1 1 1 1 0 ...
# $ HighChol            : num  1 0 1 0 1 1 0 1 1 0 ...
# $ CholCheck           : num  1 0 1 1 1 1 1 1 1 1 ...
# $ BMI                 : num  40 25 28 27 24 25 30 25 30 24 ...
# $ Smoker              : num  1 1 0 0 0 1 1 1 1 0 ...
# $ Stroke              : num  0 0 0 0 0 0 0 0 0 0 ...
# $ HeartDiseaseorAttack: num  0 0 0 0 0 0 0 0 1 0 ...
# $ PhysActivity        : num  0 1 0 1 1 1 0 1 0 0 ...
# $ Fruits              : num  0 0 1 1 1 1 0 0 1 0 ...
# $ Veggies             : num  1 0 0 1 1 1 0 1 1 1 ...
# $ HvyAlcoholConsump   : num  0 0 0 0 0 0 0 0 0 0 ...
# $ AnyHealthcare       : num  1 0 1 1 1 1 1 1 1 1 ...
# $ NoDocbcCost         : num  0 1 1 0 0 0 0 0 0 0 ...
# $ GenHlth             : num  5 3 5 2 2 2 3 3 5 2 ...
# $ MentHlth            : num  18 0 30 0 3 0 0 0 30 0 ...
# $ PhysHlth            : num  15 0 30 0 0 2 14 0 30 0 ...
# $ DiffWalk            : num  1 0 1 0 0 0 0 1 1 0 ...
# $ Sex                 : num  0 0 0 0 0 1 0 0 0 1 ...
# $ Age                 : num  9 7 9 11 11 10 9 11 9 8 ...
# $ Education           : num  4 6 4 3 5 6 6 4 5 4 ...
# $ Income              : num  3 1 8 6 4 8 7 4 1 3 ...

# Loading in the data from the proper location
data_location <- here::here("data","processed-data","processeddata.rds")

# Using the readRDS() function to load the data into a variable
diabetes <- readRDS(data_location)

# Displaying the structure
str(diabetes)
```

## SETUP - Processing Data

Although we processed the data prior, I will perform some further processing to ensure the data works properly when used in the models. Here is the structure of the dataset, so variables will be modified accordingly.

| Feature              | Description                                                                                                                                                                                                                                                                                                                             |
|------------------|------------------------------------------------------|
| ID                   | Patient ID                                                                                                                                                                                                                                                                                                                              |
| Diabetes_binary      | 0 = no diabetes 1 = prediabetes or diabetes                                                                                                                                                                                                                                                                                             |
| HighBP               | 0 = no high BP 1 = high BP                                                                                                                                                                                                                                                                                                              |
| HighChol             | 0 = no high cholesterol 1 = high cholesterol                                                                                                                                                                                                                                                                                            |
| CholCheck            | 0 = no cholesterol check in 5 years 1 = yes cholesterol check in 5 years                                                                                                                                                                                                                                                                |
| BMI                  | Body Mass Index                                                                                                                                                                                                                                                                                                                         |
| Smoker               | Have you smoked at least 100 cigarettes in your entire life? \[Note: 5 packs = 100 cigarettes\] 0 = no 1 = yes                                                                                                                                                                                                                          |
| Stoke                | (Ever told) you had a stroke. 0 = no 1 = yes                                                                                                                                                                                                                                                                                            |
| HeartDiseaseorAttack | coronary heart disease (CHD) or myocardial infarction (MI) 0 = no 1 = yes                                                                                                                                                                                                                                                               |
| PhysActivity         | physical activity in past 30 days - not including job 0 = no 1 = yes                                                                                                                                                                                                                                                                    |
| Fruits               | Consume Fruit 1 or more times per day 0 = no 1 = yes                                                                                                                                                                                                                                                                                    |
| Veggies              | Consume Vegetables 1 or more times per day 0 = no 1 = yes                                                                                                                                                                                                                                                                               |
| HvyAlchoholConsump   | Heavy drinkers (adult men having more than 14 drinks per week and adult women having more than 7 drinks per week) 0 = no 1 = yes                                                                                                                                                                                                        |
| AnyHealthcare        | Have any kind of health care coverage, including health insurance, prepaid plans such as HMO, etc. 0 = no 1 = yes                                                                                                                                                                                                                       |
| NoDocbcCost          | Was there a time in the past 12 months when you needed to see a doctor but could not because of cost? 0 = no 1 = yes                                                                                                                                                                                                                    |
| GenHlth              | Would you say that in general your health is: scale 1-5 1 = excellent 2 = very good 3 = good 4 = fair 5 = poor                                                                                                                                                                                                                          |
| MentHlth             | Now thinking about your mental health, which includes stress, depression, and problems with emotions, for how many days during the past 30 days was your mental health not good? scale 1-30 days                                                                                                                                        |
| PhysHlth             | Now thinking about your physical health, which includes physical illness and injury, for how many days during the past 30 days was your physical health not good? scale 1-30 days                                                                                                                                                       |
| DiffWalk             | Do you have serious difficulty walking or climbing stairs? 0 = no 1 = yes                                                                                                                                                                                                                                                               |
| Sex                  | 0 = female 1 = male                                                                                                                                                                                                                                                                                                                     |
| Age                  | 13-level age category (\_AGEG5YR see codebook) 1 = 18-24 9 = 60-64 13 = 80 or older                                                                                                                                                                                                                                                     |
| Education            | Education level (EDUCA see codebook) scale 1-6 1 = Never attended school or only kindergarten 2 = Grades 1 through 8 (Elementary) 3 = Grades 9 through 11 (Some high school) 4 = Grade 12 or GED (High school graduate) 5 = College 1 year to 3 years (Some college or technical school) 6 = College 4 years or more (College graduate) |
| Income               | Income scale (INCOME2 see codebook) scale 1-8 1 = less than \$10,000 5 = less than \$35,000 8 = \$75,000 or more                                                                                                                                                                                                                        |

```{r}
# This chunk is used to further process the data

# Copying data
proc_data <- diabetes

# Removing ID column
proc_data$ID <- NULL

# Target
proc_data$Diabetes_012 <- as.factor(proc_data$Diabetes_012)
levels(proc_data$Diabetes_012) <- make.names(levels(proc_data$Diabetes_012))

# Features
proc_data$HighBP <- as.factor(proc_data$HighBP)
proc_data$HighChol <- as.factor(proc_data$HighChol)
proc_data$CholCheck <- as.factor(proc_data$CholCheck)
proc_data$Smoker <- as.factor(proc_data$Smoker)
proc_data$Stroke <- as.factor(proc_data$Stroke)
proc_data$HeartDiseaseorAttack <- as.factor(proc_data$HeartDiseaseorAttack)
proc_data$PhysActivity <- as.factor(proc_data$PhysActivity)
proc_data$Fruits <- as.factor(proc_data$Fruits)
proc_data$Veggies <- as.factor(proc_data$Veggies)
proc_data$HvyAlcoholConsump <- as.factor(proc_data$HvyAlcoholConsump)
proc_data$AnyHealthcare <- as.factor(proc_data$AnyHealthcare)
proc_data$NoDocbcCost <- as.factor(proc_data$NoDocbcCost)
proc_data$GenHlth <- as.factor(proc_data$GenHlth)
proc_data$MentHlth <- as.factor(proc_data$MentHlth)
proc_data$PhysHlth <- as.factor(proc_data$PhysHlth)
proc_data$DiffWalk <- as.factor(proc_data$DiffWalk)
proc_data$Sex <- as.factor(proc_data$Sex)
proc_data$Age <- as.factor(proc_data$Age)
proc_data$Education <- as.factor(proc_data$Education)
proc_data$Income <- as.factor(proc_data$Income)

str(proc_data)
```

We can make note that a majority of our features are factors simply because most of the data recorded was on a scale, either binary (0-1), or a larger (1-8). Due to the data having so many categorical variables, we should opt to use models that more effectively handle categorical features. Therefore, we will start with a simple logistic regression, making use of one-hot encoding, then we will move on to models that may perform better such as random forest or KNN.

# Statistical Analysis

## Multinomial Logistic Regression

### One-Hot Encoding

```{r}
# Getting random sample
set.seed(123) # for reproducibility
sample_size <- 25500
sample_indices <- sample(1:nrow(proc_data), size = sample_size, replace = FALSE)
sample_data <- proc_data[sample_indices, ]
```

```{r}
# Perform one-hot encoding
sample_data_encoded <- dummy_cols(sample_data, remove_first_dummy = TRUE)

# Drop the original categorical columns
categorical_columns <- c(
  "HighBP", "HighChol", "CholCheck", "Smoker", "Stroke",
  "HeartDiseaseorAttack", "PhysActivity", "Fruits", "Veggies", "HvyAlcoholConsump",
  "AnyHealthcare", "NoDocbcCost", "GenHlth", "MentHlth", "PhysHlth", "DiffWalk",
  "Sex", "Age", "Education", "Income", "Diabetes_012_1", "Diabetes_012_2"
)

# Select only the non-categorical columns and the one-hot encoded columns
sample_data_encoded <- sample_data_encoded[, !names(sample_data_encoded) %in% categorical_columns]

# Rename the factor levels to valid variable names
levels(sample_data_encoded$Diabetes_012) <- make.names(levels(sample_data_encoded$Diabetes_012))

# Inspect the new data frame
str(sample_data_encoded)
```

### Training

```{r}
set.seed(12345)
train_control <- trainControl(method = "cv", number = 5, 
                              classProbs = TRUE, 
                              summaryFunction = multiClassSummary)
```

```{r}
set.seed(12345)
logistic_model <- train(Diabetes_012 ~ ., 
                        data = sample_data_encoded, 
                        method = "multinom", 
                        trControl = train_control,
                        preProcess = c("center", "scale"),
                        metric = "Accuracy")  # Optimization metric can be "Accuracy" or "Kappa"

```

### Results

```{r}
print(logistic_model)
```

```{r}
# Make predictions on the same data (or a test dataset if you have one)
predictions <- predict(logistic_model, newdata = sample_data_encoded)

# Confusion matrix
confusion_matrix <- confusionMatrix(predictions, sample_data_encoded$Diabetes_012)
print(confusion_matrix)
```

## Random Forest

### Sampling Data

I may need to change how I am sampling the data so I can make sure the sample is balanced.

```{r}
# Getting random sample
set.seed(123) # for reproducibility
sample_size <- 25500
sample_indices <- sample(1:nrow(proc_data), size = sample_size, replace = FALSE)
sample_data <- proc_data[sample_indices, ]
```

### Training

```{r}
set.seed(12345)
train_control <- trainControl(method = "cv", number = 2, 
                              classProbs = TRUE, 
                              summaryFunction = multiClassSummary)

# Fit the random forest model using caret
set.seed(12345)
rf_model <- train(Diabetes_012 ~ ., 
                  data = sample_data, 
                  method = "rf", 
                  trControl = train_control,
                  preProcess = c("center", "scale"),
                  metric = "Accuracy")  # Optimization metric can be "Accuracy" or "Kappa"
```

### Results

```{r}
print(rf_model)
```

```{r}
# Make predictions on the same data (or a test dataset if you have one)
predictions <- predict(rf_model, newdata = sample_data)

# Confusion matrix
confusion_matrix <- confusionMatrix(predictions, sample_data$Diabetes_012)
print(confusion_matrix)
```

## KNN

### Training

```{r}

```

### Results

```{r}

```

### 2024-06-22 code/eda-code/EDA-Code.qmd
Updated Statistical Analysis Code - Final Project Part 3
Worked on the statistical analysis code for part 3 of the final project.

Updated manuscript with new relevant information from statistical analysis process.

Updated EDA code to create histograms all in one image instead of seperate images.

@@
 library(GGally) # Pais ggplot2
+library(gridExtra)
 ```
@@
 
-```{r}
-
-```
-
 # Plots of Response/Features
@@
 ```{r}
-p1 <- mydata %>% ggplot(aes(x=Diabetes_012)) + geom_histogram() 
-plot(p1)
-figure_file = here("results", "figures","diabetes-distribution.png")
-ggsave(filename = figure_file, plot=p1) 
-```
+# Function to create histograms for each variable
+create_histograms <- function(data) {
+  plots <- list()
+  for (col_name in names(data)) {
+    if (is.numeric(data[[col_name]])) {
+      p <- ggplot(data, aes_string(x = col_name)) + 
+        geom_histogram(binwidth = 1, fill = "salmon", color = "black") +
+        labs(title = col_name) +
+        theme_minimal()
+    } else {
+      p <- ggplot(data, aes_string(x = col_name)) + 
+        geom_bar(fill = "salmon", color = "black") +
+        labs(title = col_name) +
+        theme_minimal()
+    }
+    plots[[col_name]] <- p
+  }
+  return(plots)
+}
 
-```{r}
-p2 <- mydata %>% ggplot(aes(x=Smoker)) + geom_histogram() 
-plot(p2)
-figure_file = here("results", "figures","smoker-distribution.png")
-ggsave(filename = figure_file, plot=p2) 
-```
+# Create histograms for all variables
+histograms <- create_histograms(mydata)
 
-```{r}
-p3 <- mydata %>% ggplot(aes(x=BMI)) + geom_histogram() 
-plot(p3)
-figure_file = here("results", "figures","BMI-distribution.png")
-ggsave(filename = figure_file, plot=p3) 
-```
+# Arrange the plots in a grid
+n <- length(histograms)
+nCol <- ceiling(sqrt(n))
+nRow <- ceiling(n / nCol)
+combined_plot <- marrangeGrob(grobs = histograms, nrow = nRow, ncol = nCol)
 
-```{r}
-p4 <- mydata %>% ggplot(aes(x=PhysHlth)) + geom_histogram() 
-plot(p4)
-figure_file = here("results", "figures","PhysHlth-distribution.png")
-ggsave(filename = figure_file, plot=p4) 
-```
-
-```{r}
-p5 <- mydata %>% ggplot(aes(x=Age)) + geom_histogram() 
-plot(p5)
-figure_file = here("results", "figures","Age-distribution.png")
-ggsave(filename = figure_file, plot=p5) 
+# Save the combined plot to a PNG file
+figure_file <- here("results", "figures", "variable-distributions.png")
+ggsave(filename = figure_file, plot = combined_plot, width = 20, height = 20)
 ```

### 2024-06-22 products/manuscript/Manuscript.qmd
Updated Statistical Analysis Code - Final Project Part 3
Worked on the statistical analysis code for part 3 of the final project.

Updated manuscript with new relevant information from statistical analysis process.

Updated EDA code to create histograms all in one image instead of seperate images.

@@
 | Feature              | Description                                                                                                                                                                                                                                                                                                                             |
-|------------|------------------------------------------------------------|
+|------------------|------------------------------------------------------|
 | ID                   | Patient ID                                                                                                                                                                                                                                                                                                                              |
@@
 
-*\*To be added in later portion of project*
+Not much occurs during the cleaning stage as the dataset was processed and cleaned prior.
+The only modifications made to the dataset was removing the 'ID' column as it is not useful for this analysis.
+Once the data was imported, it was only a matter of converting categorical variables to factors so they would function properly.
+One-hot encoding was implemented, but only for the multinomial logistic regression.
 
-## Statistical analysis
+## Statistical analysis method
 
@@
 After viewing the correlations, I will then use simple linear regression and random forest models to identify variables of importance, these variables will be noted and used later during the modeling phase.
-I will then split the data set into a training and test set using a 80/20 split and begin the modeling phase.
+I will utilize cross validation during the modeling phase.
 This section will be filled out more once we are further into the class and have discussed more approaches/methods.
-
-***\* Any section past results has not been modified yet. I only completed the sections required for the first part.***
 
@@
 
-### Diabetes Distribution
+### Distribution of all variables
 
 ```{r}
-knitr::include_graphics(here("results","figures","diabetes-distribution.png"))
-```
-
-### BMI Distribution
-
-```{r}
-knitr::include_graphics(here("results","figures","BMI-distribution.png"))
-```
-
-### PhysHlth Distribution
-
-```{r}
-knitr::include_graphics(here("results","figures","PhysHlth-distribution.png"))
-```
-
-### Age Distribution
-
-```{r}
-knitr::include_graphics(here("results","figures","Age-distribution.png"))
-```
-
-### Smoker Distribution
-
-```{r}
-knitr::include_graphics(here("results","figures","smoker-distribution.png"))
+knitr::include_graphics(here("results","figures","variable-distributions.png"))
 ```
@@
 
-***\*To be added in later portion of project***
+***\*IN PROGRESS***
+
+### Methods
+
+After processing and exploring the data, it was discovered that a majority of the features in this dataset are categorical.
+Therefore, we will be making use of models that are more rigorous to our situation, such as random forest and KNN.
+However before trying those models, we will first do a simple multinomial logistic regression.
+Why multinomial?
+That is because our response variable is not binary, it has three levels to it; typically, logistic regression is used for binary classification.
+After creating models with all features included, we will go on to look at models with features that fit our importance criteria (i.e.
+look at the models results when using the most important features).
+
+### Preparing the data
+
+For some of the models, mainly multinomial logistic regression, we need to further process our data.
+The method we will use is known as one-hot encoding.
+One-hot encoding is a process of converting categorical variables into a binary (0 or 1) format where each category becomes a new binary column.
+We need to do this because most machine learning algorithms require numerical input and cannot handle categorical data directly, so formatting this data in a different manner will provide us with better results.
+
+### Results
+
+#### Multinomial Logistic Regression - All Features
+
+```         
+Confusion Matrix and Statistics
+
+          Reference
+Prediction    X0    X1    X2
+        X0 21075   398  2891
+        X1     3     1     1
+        X2   477    32   622
+
+Overall Statistics
+
+               Accuracy : 0.8509          
+                 95% CI : (0.8465, 0.8553)
+    No Information Rate : 0.8453          
+    P-Value [Acc > NIR] : 0.006606
+
+                  Kappa : 0.1995
+
+ Mcnemar's Test P-Value : < 2.2e-16
+
+Statistics by Class:
+
+                     Class: X0 Class: X1 Class: X2
+Sensitivity             0.9777 2.320e-03   0.17701
+Specificity             0.1663 9.998e-01   0.97685
+Pos Pred Value          0.8650 2.000e-01   0.54996
+Neg Pred Value          0.5775 9.831e-01   0.88132
+Prevalence              0.8453 1.690e-02   0.13780
+Detection Rate          0.8265 3.922e-05   0.02439
+Detection Prevalence    0.9555 1.961e-04   0.04435
+Balanced Accuracy       0.5720 5.011e-01   0.57693
+```
+
+\- Notes (Will be expanded and explained in future update):
+
+-   Accuracy: 85.09%
+
+-   Target variable may not be balanced
+
+#### Random Forest - All Features
+
+```         
+Confusion Matrix and Statistics
+
+          Reference
+Prediction    X0    X1    X2
+        X0 21555   431  3514
+        X1     0     0     0
+        X2     0     0     0
+
+Overall Statistics
+
+               Accuracy : 0.8453          
+                 95% CI : (0.8408, 0.8497)
+    No Information Rate : 0.8453          
+    P-Value [Acc > NIR] : 0.5042
+
+                  Kappa : 0
+
+ Mcnemar's Test P-Value : NA
+
+Statistics by Class:
+
+                     Class: X0 Class: X1 Class: X2
+Sensitivity             1.0000    0.0000    0.0000
+Specificity             0.0000    1.0000    1.0000
+Pos Pred Value          0.8453       NaN       NaN
+Neg Pred Value             NaN    0.9831    0.8622
+Prevalence              0.8453    0.0169    0.1378
+Detection Rate          0.8453    0.0000    0.0000
+Detection Prevalence    1.0000    0.0000    0.0000
+Balanced Accuracy       0.5000    0.5000    0.5000
+```
+
+\- Notes (Will be expanded and explained in future update):
+
+-   Model is only predicting '0', likely an error in model setup
+
+-   Due to inaccurate predictions, accuracy is not applicable.
+
+#### KNN - All Features
+
+-   in progress
 
@@
 
-***\*To be added in later portion of project***
+***\*To be updated in later portion of project***
 
@@
 
-***\*To be added in later portion of project***
+***\*To be updated in later portion of project***
+
+-   LIMIT: Having a lot of categorical features
+
+-   LIMIT: Having a response variable with 3 levels instead of a binary.
 
@@
 
-***\*To be added in later portion of project***
+***\*To be updated in later portion of project***

### 2024-07-13 code/analysis-code/statistical_analysis_code.qmd
Finished Part 4
Completed the modeling and updated the manuscript, currently working on references

@@
 library(randomForest)
+library(here) #to set paths
 ```
@@
 | Feature              | Description                                                                                                                                                                                                                                                                                                                             |
-|------------------|------------------------------------------------------|
+|------------|------------------------------------------------------------|
 | ID                   | Patient ID                                                                                                                                                                                                                                                                                                                              |
@@
 ```{r}
-
+set.seed(12345)
+knn_model <- train(Diabetes_012 ~ ., 
+                   data = sample_data, 
+                   method = "knn", 
+                   trControl = train_control,
+                   preProcess = c("center", "scale"),
+                   metric = "Accuracy") 
 ```
@@
 ```{r}
+# Make predictions on the same data (or a test dataset if you have one)
+predictions <- predict(knn_model, newdata = sample_data)
 
+# Confusion matrix
+confusion_matrix <- confusionMatrix(predictions, sample_data$Diabetes_012)
+print(confusion_matrix)
 ```
+
+Plot of accuracy
+
+```{r}
+accuracy <- data.frame(
+  Model = c("Logistic Regression", "Random Forest", "KNN"),
+  Accuracy = c(
+    max(logistic_model$results$Accuracy),
+    max(rf_model$results$Accuracy),
+    max(knn_model$results$Accuracy)
+  )
+)
+
+accuracy_plot <- ggplot(accuracy, aes(x = Model, y = Accuracy, fill = Model)) +
+  geom_bar(stat = "identity") +
+  theme_minimal() +
+  labs(title = "Model Accuracy Comparison", y = "Accuracy", x = "Model") +
+  scale_fill_brewer(palette = "Set2")
+```
+
+```{r}
+# Define the file path to save the plot
+figure_file <- here("results", "figures")
+
+ggsave(filename = "model_accuracy_comparison.png", plot = accuracy_plot, path = figure_file)
+```

### 2024-07-13 products/manuscript/Manuscript.qmd
Finished Part 4
Completed the modeling and updated the manuscript, currently working on references

@@
 | Feature              | Description                                                                                                                                                                                                                                                                                                                             |
-|------------------|------------------------------------------------------|
+|--------------------|----------------------------------------------------|
 | ID                   | Patient ID                                                                                                                                                                                                                                                                                                                              |
@@
 
-*\*Analysis methods will be included in later part of project*
+The main question I would like to answer with this project is "What are the key health indicators that predict the likelihood of an individual being diagnosed with diabetes?".
 
@@
 
-### Explain the question to be answered using the data
+This data comes from CDC and will be sampled during analysis.
 
-The main question I would like to answer with this project is "What are the key health indicators that predict the likelihood of an individual being diagnosed with diabetes?".
-
-### What will be your outcome(s) of interest (if any)?
+### Outcome of Interest
 
@@
 For my analysis, I will first look at the distributions of each variable to identify if the dataset is balanced and outliers.
-I do not think any data cleaning will need to be done as that was completed when the data was collected from the CDCs website.
+I do not think any data cleaning will need to be done as that was completed when the data was collected from the CDC's website.
 Next, I will create a correlation matrix to view the correlations between each predictor and the response variable.
@@
 This section will be filled out more once we are further into the class and have discussed more approaches/methods.
+
+#### Logistic Regression
+
+Logistic regression will be used to model the relationship between the binary outcome variable and the predictors.
+This method will help in understanding the influence of each predictor on the likelihood of diabetes diagnosis.
+
+#### Random Forest
+
+Random forest, an ensemble learning method, will be used to improve the accuracy and robustness of predictions.
+It will help in identifying the importance of each predictor in determining the diabetes diagnosis.
+
+#### K-Nearest Neighbors (KNN)
+
+KNN will be used as a non-parametric method to classify individuals based on the closest training examples in the feature space.
+This will provide insights into the influence of neighborhood-based predictions on diabetes diagnosis.
 
@@
 
-\- Notes (Will be expanded and explained in future update):
+\- Notes (Will be expanded and explained in next update):
 
@@
 -   in progress
+
+#### Model Comparisons
+
+```{r}
+knitr::include_graphics(here("results","figures","model_accuracy_comparison.png"))
+```
 

### 2024-07-27 products/manuscript/Manuscript.qmd
Completed Manuscript
Finalized the KNN model and then compared all three. Outlined the results and provided a summary as well as limitations. Once analysis was complete, wrote an abstract and completed readme files.

@@
     highlight-style: github
+bibliography: ../../assets/dataanalysis-references.bib
 csl: ../../assets/apa.csl
@@
 
-*\*Summary will be finalized in later part of project and will include setup and results*
+Diabetes is a chronic condition that affects millions of people worldwide, characterized by high blood glucose levels and serious complications if not managed properly.
+This study aims to identify key health indicators predicting the likelihood of an individual being diagnosed with diabetes using a dataset from the UC Irvine Machine Learning Repository, containing over 250,000 instances and 21 features.
+Three models were utilized: logistic regression, random forest, and K-nearest neighbors (KNN).
+The analysis revealed that BMI, high blood pressure, heart disease or attack history, and heavy alcohol consumption were significant predictors of diabetes.
+Logistic regression and KNN models demonstrated reasonable accuracy, while the random forest model required refinement.
+The study highlights the importance of these health indicators and suggests further data processing and model tuning for improved predictive performance.
 
@@
 
-*\*more background information will be added later, I will describe what diabetes is, what are the notable influences on diagnosis, and what the dataset provides in relation to other studies on diabetes diagnosis.*
+"Diabetes can lead to severe, and in some cases, life-threatening complications." [@ADA].
+Diabetes is a major cause of blindness, kidney failure, heart attacks, stroke, and lower limb amputation.
+Managing diabetes and its complications places a significant burden on healthcare systems worldwide.
+Early diagnosis and management are crucial to prevent adverse health outcomes.
+Several studies have identified various risk factors associated with diabetes, including age, BMI, physical activity, and dietary habits.
 
@@
 | Feature              | Description                                                                                                                                                                                                                                                                                                                             |
-|--------------------|----------------------------------------------------|
+|---------------------|---------------------------------------------------|
 | ID                   | Patient ID                                                                                                                                                                                                                                                                                                                              |
@@
 
+For my model selection, I referenced ISLR2. [@ISLR]
+
 {{< pagebreak >}}
@@
 
-\*Will fix and put into readable table
-
-```         
-            HighBP             HighChol            CholCheck                  BMI               Smoker               Stroke HeartDiseaseorAttack 
-            1.141777             1.070725             1.011236             1.124242             1.073702             1.071638             1.150480 
-        PhysActivity               Fruits              Veggies    HvyAlcoholConsump        AnyHealthcare          NoDocbcCost              GenHlth 
-            1.140832             1.109908             1.105040             1.013591             1.100212             1.157580             1.743109 
-            MentHlth             PhysHlth             DiffWalk                  Sex                  Age            Education               Income 
-            1.281663             1.697022             1.499052             1.109269             1.326329             1.299898             1.505502 
+```{r}
+knitr::include_graphics(here("results","figures","importance_table.png"))
 ```
@@
 ## Basic statistical analysis
-
-***\*IN PROGRESS***
 
@@
 
-\- Notes (Will be expanded and explained in next update):
-
--   Accuracy: 85.09%
-
--   Target variable may not be balanced
+The logistic regression model was applied to the dataset to analyze the relationship between the binary outcome variable (diabetes diagnosis) and the predictors.
+The model achieved an accuracy of 85.09%.
+However, the model showed that the target variable might not be balanced, indicating that further data processing might be necessary.
 
@@
 
-\- Notes (Will be expanded and explained in next update):
-
--   Model is only predicting '0', likely an error in model setup
-
--   Due to inaccurate predictions, accuracy is not applicable.
+The random forest model was applied to the dataset to improve the accuracy and robustness of predictions.
+The model, however, only predicted the class '0', likely due to an error in model setup.
+This resulted in an accuracy that was not applicable, indicating the need for further refinement of the model.
 
@@
 
--   in progress
+```         
+Confusion Matrix and Statistics
+
+          Reference
+Prediction    X0    X1    X2
+        X0 21270   415  2975
+        X1     0     0     0
+        X2   285    16   539
+
+Overall Statistics
+
+               Accuracy : 0.8553          
+                 95% CI : (0.8509, 0.8596)
+    No Information Rate : 0.8453          
+    P-Value [Acc > NIR] : 4.769e-06
+
+                  Kappa : 0.1869
+
+ Mcnemar's Test P-Value : < 2.2e-16
+
+Statistics by Class:
+
+                     Class: X0 Class: X1 Class: X2
+Sensitivity             0.9868    0.0000   0.15339
+Specificity             0.1407    1.0000   0.98631
+Pos Pred Value          0.8625       NaN   0.64167
+Neg Pred Value          0.6607    0.9831   0.87936
+Prevalence              0.8453    0.0169   0.13780
+Detection Rate          0.8341    0.0000   0.02114
+Detection Prevalence    0.9671    0.0000   0.03294
+Balanced Accuracy       0.5637    0.5000   0.56985
+```
+
+The KNN model was applied to the dataset and achieved an accuracy of 85.53%.
+The model's sensitivity and specificity values indicated that it performed reasonably well, but there was room for improvement, particularly in predicting minority classes.
 
@@
 
-## Full analysis
-
-***\*To be updated in later portion of project***
-
 {{< pagebreak >}}
@@
 
-***\*To be updated in later portion of project***
+The analysis revealed that BMI, HighBP, HeartDiseaseorAttack, and HvyAlchoholConsump were significant predictors of diabetes diagnosis.
+Logistic regression and KNN models showed reasonable accuracy, while the random forest model required further refinement.
+These findings highlight the importance of these health indicators in predicting diabetes and suggest that further data processing and model tuning could improve predictive performance.
 
@@
 
-***\*To be updated in later portion of project***
+-   **Imbalanced Dataset**: The dataset was imbalanced, which affected the model's performance, particularly in predicting minority classes.
 
--   LIMIT: Having a lot of categorical features
+-   **Categorical Data Handling**: The handling of categorical variables posed challenges, especially for models like logistic regression and random forest.
 
--   LIMIT: Having a response variable with 3 levels instead of a binary.
+-   **Model Setup Errors**: The random forest model's performance indicated potential errors in model setup, necessitating further refinement.
 
@@
 
-***\*To be updated in later portion of project***
+The study identified key health indicators that are strongly associated with diabetes diagnosis.
+Logistic regression and KNN models provided reasonable predictive accuracy, while the random forest model required further refinement.
+These findings underscore the need for balanced datasets and appropriate handling of categorical variables to improve model performance.
 
@@
 
-***\*To be updated in later portion of project***
+For some reason my bib file was being weird, will try to fix.

### 2024-08-08 products/manuscript/Manuscript.qmd
Final Project Submission
Fixed bibtex and added a few more things to the manuscript. Added the figures as images since knit was being strange about loading images.

@@
 This data has been gathered and processed from the CDCs website in 2014.
-This dataset includes over 250,000 instances (observations) and 21 features, ranging from personal information (income/education) to health history (drinking/smoking/physical health).
+This dataset includes over 250,000 instances (observations) and 21 features, ranging from personal information (income/education) to health history (drinking/smoking/physical health)[@UCI].
 Bellow are the features and there respective descriptions
@@
 | Feature              | Description                                                                                                                                                                                                                                                                                                                             |
-|---------------------|---------------------------------------------------|
+|----------------------|--------------------------------------------------|
 | ID                   | Patient ID                                                                                                                                                                                                                                                                                                                              |
@@
 
+![](images/variable-distributions.png){fig-align="center"}
+
 ```{r}
-knitr::include_graphics(here("results","figures","variable-distributions.png"))
+#knitr::include_graphics(here("results","figures","variable-distributions.png"))
 ```
@@
 
+![](images/correlation_matrix.png){fig-align="center"}
+
 ```{r}
-knitr::include_graphics(here("results","figures","correlation_matrix.png"))
+#knitr::include_graphics(here("results","figures","correlation_matrix.png"))
 ```
@@
 
+![](images/importance_table.png){fig-align="center" width="300"}
+
 ```{r}
-knitr::include_graphics(here("results","figures","importance_table.png"))
+#knitr::include_graphics(here("results","figures","importance_table.png"))
 ```
@@
 
+![](images/variable_importance.png){fig-align="center"}
+
 ```{r}
-knitr::include_graphics(here("results","figures","variable_importance.png"))
+#knitr::include_graphics(here("results","figures","variable_importance.png"))
 ```
@@
 
+![](images/model_accuracy_comparison.png){fig-align="center"}
+
 ```{r}
-knitr::include_graphics(here("results","figures","model_accuracy_comparison.png"))
+#knitr::include_graphics(here("results","figures","model_accuracy_comparison.png"))
 ```
@@
 ## Strengths and Limitations
+
+**Strengths:**
+
+-   **Large Dataset**: The study utilized a large dataset with over 250,000 observations, providing a large sample size for reliable statistical analysis and model training.
+
+-   **Comprehensive Features**: The dataset had 21 diverse features ranging from personal information to health history, allowing for a thorough analysis of various health indicators.
+
+-   **Multiple Models**: The use of three different models (logistic regression, random forest, and KNN) provided a comprehensive approach to identifying key predictors and comparing model performances.
+
+**Limitations:**
 
@@
 # References
-
-For some reason my bib file was being weird, will try to fix.

### 2024-08-08 products/manuscript/Manuscript.qmd
Final Project as PDF
added a pdf version instead of html

@@
 format:
-  html:
+  pdf:
     toc: true
